Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-10-23 TACO-paper-v2/coffee.tex
changes suggested by Paul on Oct 22 via email

\begin{abstract}
We study and systematically evaluate a class of composable code
transformations that improve arithmetic intensity in local assembly
operations, which represent a significant fraction of the execution
time in finite element methods. Their performance optimization is indeed a
challenging issue. Even though affine loop nests are generally present,
the short trip counts and the complexity of mathematical expressions, which
vary among different problems, make it hard to determine a single or unique
sequence of successful transformations. Our investigation has resulted in the
implementation of a compiler for local assembly kernels, fully
integrated with a framework for developing finite element methods.
The compiler manipulates abstract syntax trees generated from a domain-specific
language by introducing domain-aware optimizations for instruction-level parallelism and register locality. Eventually,
it produces C code including vector SIMD intrinsics. Experiments using a range of real-world finite-element
problems of increasing complexity show that significant performance improvement
is achieved. The generality of the approach and the applicability of the proposed
code transformations to other domains is also discussed.

@@

Achieving high performance on CPUs is non-trivial. The complexity of the
mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine a
single or specific sequence of transformations that is successfully
applicable to different problems (equations). Loop trip counts are typically small and can
vary significantly, which further exacerbates the issue. A
compiler-based approach is, therefore, the only reasonable option to
obtain close-to-peak performance in a wide range of different local
assembly kernels. Optimizations like padding, generalized loop-invariant
code motion, vector-register tiling, and expression splitting, as well
as their composition, are essential, but their support in
state-of-the-art polyhedral and vendor compilers, if any, is very
limited. BLAS routines could theoretically be employed, although a
fairly complicated control and data flow analysis would be required to
automate identification and extraction of matrix-matrix multiplies. In
addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
dimension of the matrices involved and the potential loss in data
locality can limit or eliminate the performance gain of this approach.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, has short trip counts (in the range 3--15), and the computation
reduces to a summation of a few products involving basis functions. An
example is provided in Listing~\ref{code:helmholtz}, which shows the
assembly kernel for a Helmholtz problem using Lagrange basis functions
on 2D elements with polynomial order $p=1$. In other scenarios, for
instance when solving the Burgers equation, the number of arrays
involved in the computation of the element matrix can be much larger.
The assembly code is given in Listing~\ref{code:burgers} and contains 14
unique arrays that are accessed, where the same array can be referenced
multiple times within the same expression. This may also require the
evaluation of constants in outer loops (called $F$ in the code) to act
as scaling factors of arrays. Trip counts grow proportionately to the
order of the method and arrays may be block-sparse. In addition to a
larger number of operations, more complex cases like the Burgers
equation are characterized by high register pressure.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for
improving arithmetic intensity in finite element local assembly
kernels, and their integration in a novel compiler, COFFEE. In the
context of automated code generation for finite element local
assembly, COFFEE is the first compiler capable of introducing
low-level optimizations to maximize instruction-level parallelism,
register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very
small, with the size depending on aspects like the degree of accuracy
one wants to reach (polynomial order of the method) and the mesh
discretization employed.  The data space, in terms of number of arrays
and scalars required to evaluate the element matrix, grows
proportionately with the complexity of the finite element
problem. COFFEE has been developed taking into account all of these
degrees of freedom, based on the idea that reducing the problem of
local assembly optimization to a fixed sequence of transformations is
far too superficial if close-to-peak performance needs to be
reached. The various optimizations overcome limitations of current
vendor and research compilers. The exploitation of domain knowledge
allows some of them to be particularly effective, as demonstrated by
our experiments on two state-of-the-art Intel platforms. Further work
includes a comprehensive study about feasibility and constraints on
transforming kernels into a sequence of calls to external linear
algebra libraries. As suggested by the performance evaluation, an
exploration of more sophisticated approaches for expression splitting
also looks promising. COFFEE supports all of the problems expressible
in Firedrake, and is already integrated with this framework. The
generality and the applicability of the proposed code transformations
to other domains has also been discussed.

>>>>

\begin{abstract}
We study and systematically evaluate a class of composable code
transformations that improve arithmetic intensity in local assembly
operations, which represent a significant fraction of the execution
time in finite element methods. Their performance optimization is
indeed a challenging issue. Even though affine loop nests are
generally present, the short trip counts and the complexity of
mathematical expressions, which vary among different problems, make it
hard to determine an optimal sequence of successful
transformations. Our investigation has resulted in the implementation
of a compiler for local assembly kernels, fully integrated with a
framework for developing finite element methods.  The compiler
manipulates abstract syntax trees generated from a domain-specific
language by introducing domain-aware optimizations for
instruction-level parallelism and register locality. Eventually, it
produces C code including vector SIMD intrinsics. Experiments using a
range of real-world finite-element problems of increasing complexity
show that significant performance improvement is achieved. The
generality of the approach and the applicability of the proposed code
transformations to other domains is also discussed.

@@

Achieving high performance on CPUs is non-trivial. The complexity of
the mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine
an optimal sequence of transformations that is successfully applicable
to different problems (equations). Loop trip counts are typically
small and can vary significantly, which further exacerbates the
issue. A compiler-based approach is, therefore, the only reasonable
option to obtain close-to-peak performance in a wide range of
different local assembly kernels. Optimizations like padding,
generalized loop-invariant code motion, vector-register tiling, and
expression splitting, as well as their composition, are essential, but
their support in state-of-the-art polyhedral and vendor compilers, if
any, is very limited. BLAS routines could theoretically be employed,
although a fairly complicated control and data flow analysis would be
required to automate identification and extraction of matrix-matrix
multiplies. In addition, as detailed in
Section~\ref{sec:perf-eval-blas}, the small dimension of the matrices
involved and the potential loss in data locality can limit or
eliminate the performance gain of this approach.

@@

The actual complexity of a local assembly kernel depends on the finite
element problem being solved. In simpler cases, the loop nest is
perfect, has short trip counts (in the range 3--15), and the computation
reduces to a summation of a few products involving basis functions. An
example is provided in Listing~\ref{code:helmholtz}, which shows the
assembly kernel for a Helmholtz problem using Lagrange basis functions
on 2D elements with polynomial order $p=1$. In other scenarios, for
instance when solving the Burgers equation, the number of arrays
involved in the computation of the element matrix can be much larger.
The assembly code is given in Listing~\ref{code:burgers} and contains 14
different arrays that are accessed, where the same array can be referenced
multiple times within the same expression. This may also require the
evaluation of constants in outer loops (called $F$ in the code) to act
as scaling factors of arrays. Trip counts grow proportionately to the
order of the method and arrays may be block-sparse. In addition to a
larger number of operations, more complex cases like the Burgers
equation are characterized by high register pressure.

@@

In this paper, we have presented the study and systematic performance
evaluation of a class of composable cross-loop optimizations for
improving arithmetic intensity in finite element local assembly
kernels, and their integration in a novel compiler, COFFEE. In the
context of automated code generation for finite element local
assembly, COFFEE is the first compiler capable of introducing
low-level optimizations to maximize instruction-level parallelism,
register locality and SIMD vectorization. Assembly kernels have
particular characteristics. Their iteration space is usually very
small, with the size depending on aspects like the degree of accuracy
one wants to reach (polynomial order of the method) and the mesh
discretization employed.  The data space, in terms of number of arrays
and scalars required to evaluate the element matrix, grows
proportionately with the complexity of the finite element
problem. COFFEE has been developed taking into account all of these
degrees of freedom, based on the idea that reducing the problem of
local assembly optimization to a fixed sequence of transformations is
far too simple-minded if close-to-peak performance needs to be
reached. The various optimizations overcome limitations of current
vendor and research compilers. The exploitation of domain knowledge
allows some of them to be particularly effective, as demonstrated by
our experiments on two state-of-the-art Intel platforms. Further work
includes a comprehensive study about feasibility and constraints on
transforming kernels into a sequence of calls to external linear
algebra libraries. As suggested by the performance evaluation, an
exploration of more sophisticated approaches for expression splitting
also looks promising. COFFEE supports all of the problems expressible
in Firedrake, and is already integrated with this framework. The
generality and the applicability of the proposed code transformations
to other domains has also been discussed.
